---
title: "INFOF422 exercise."
subtitle: "Linear regression: bias of empirical risk"
author: "G. Bontempi"
#date: "3/4/2021"
output: pdf_document

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question

Let us consider the dependency where the conditional distribution of ${\mathbf y}$ is
$$
{\mathbf y}= 1-x+x^2-x^3+{\mathbf w}
$$
where ${\mathbf w}\sim N(0,\sigma^2)$, $x \in \Re$
takes the values ${\tt seq(-1,1,length.out=N)}$ (with $N=50$) 
and $\sigma=0.5$.

Consider the family of regression models
$$h^{(m)}(x)=\beta_0 +\sum_{j=1}^m \beta_j x^j$$
where $p$ denote the number of weights of the polynomial model $h^{(m)}$ of degree $m$.

Let $\widehat{\text{MISE}}_{\text{emp}}^{(m)}$ denote the least-squares empirical risk
and MISE the mean integrated empirical risk.


By using Monte Carlo simulation and for $m=0,\dots,6$

*  plot
$E[\widehat{\textbf{MISE}}_{\text{emp}}^{(m)}]$ as a function of $p$,
* plot $\text{MISE}^{(m)}$ as a function of $p$,
* plot the difference $E[\widehat{\textbf{MISE}}_{\text{emp}}^{(m)}]-\text{MISE}^{(m)}$ 
as a function of $p$ and compare it with the theoretical result seen during the class.


For a single observed dataset:

* plot
$\widehat{\text{MISE}}_{\text{emp}}^{(m)}$ as a function of the number of model parameters $p$,
* plot PSE as a function of $p$,
* discuss the relation between $\arg \min_m \widehat{\text{MISE}}_{\text{emp}}^{(m)}$ and $\arg \min_m \text{PSE}^{(m)}$



NOTA BENE: the use of the R command $\texttt{lm}$ is NOT allowed.

\pagebreak

## Monte Carlo Simulation


```{r run MC, run=FALSE}

N=50 ## number of samples

S=10000 ## number of MC trials
M=6  ## max order of the polynomial model
sdw=0.5 ## stanard deviation of noise

Emp<-array(NA,c(M+1,S))
MISE<-array(NA,c(M+1,S))
for (s in 1:S){
  X=seq(-1,1,length.out=N)
  Y=1-X+X^2-X^3+rnorm(N,sd=sdw)
  
  Xts=X
  Yts=1-Xts+Xts^2-Xts^3+rnorm(N,sd=sdw)
  
  for (m in 0:M){
    DX=NULL
    for (j in 0:m){
      DX=cbind(DX,X^j)
    }
    
    betahat=solve(t(DX)%*%DX)%*%t(DX)%*%Y
    Yhat=DX%*%betahat
    Emp[m+1,s]=mean((Y-Yhat)^2)
    MISE[m+1,s]=mean((Yts-Yhat)^2)
   
  }
  
  
  
}
  
  mMISE=apply(MISE,1,mean)
  mEmp=apply(Emp,1,mean)
 
 

```



### Plot expected empirical risk and MISE

```{r}
plot(mMISE, ylim=c(0,1), type="l",xlab="p")

lines(mEmp,col="red")
legend(x=4,y=1,c("MISE","Empirical risk"),lty=1, col=c("black","red"))
 
```

### Plot bias of empirical risk vs theoretical quantity

```{r}
plot(mMISE-mEmp,  type="l",xlab="p")
p=1:(M+1)
lines(p,2*p*sdw^2/N,col="red")
legend(x=4,y=0.02,c("Monte Carlo bias","Theoretical bias"),lty=1, col=c("black","red"))
```

## Single dataset

```{r}
set.seed(0)
N=50 ## number of samples

M=6  ## max order of the polynomial model
sdw=0.5 ## stanard deviation of noise

Emp<-numeric(M+1)
FPE<-numeric(M+1)

X=seq(-1,1,length.out=N)
Y=1-X+X^2-X^3+rnorm(N,sd=sdw)
  

  
  for (m in 0:M){
    DX=NULL
    for (j in 0:m){
      DX=cbind(DX,X^j)
    }
    
    betahat=solve(t(DX)%*%DX)%*%t(DX)%*%Y
    Yhat=DX%*%betahat
    Emp[m+1]=mean((Y-Yhat)^2)
    sdw=sd(Y-Yhat)
    FPE[m+1]=Emp[m+1]+2*sdw^2/N*(m+1)
   
  }
  
bestEmp=which.min(Emp)-1
bestFPE=which.min(FPE)-1
  
print(bestFPE)
plot(Emp,type="l")
lines(FPE,col="red")
 
legend(x=4,y=1,c("Empirical risk","FPE"),lty=1, col=c("black","red"))

```


The  model degree **`r bestEmp`** returned by minimizing the empirical risk corresponds to the
highest order considered.

The  model degree **`r bestFPE`** returned by minimizing the empirical risk corresponds to the real degree of the regression function $E[{\mathbf y}| x]$.
